Explaining probabilistic prediction on the simplex with Shapley compositions

### 2023-12-13 paper/main.tex
edit paper

\subsection{The isometric log-ratio transformation}

@@

\subsection{Angles, norms and projections}

@@

@@

@@

\subsubsection{More classes}

\subsubsection{Histograms}

@@

In this work, the estimation algorithm we used to compute the Shapley compositions is an adaptation of Algorithm 2 in \cite{vstrumbelj2014explaining}. Since the resulting Shapley compositions are approximations, the efficiency property does not necessarily hold without an adjustement. Each Shapley compositions are therefore adjusted following a similar method as in the sampling approximation in the SHAP toolkit \cite{NIPS2017_7062}\footnote{\url{https://github.com/shap/shap/blob/master/shap/explainers/_sampling.py}}. See Appendix \ref{app:algo} and \ref{app:correct} for more details.

@@

@@

\section{Class compositions}

A $k$-class compositions $\bm{c}^{(k)}$ is defined as an unit norm composition going straight to the direction of the $k$th class. This is a discrete probability distribution with maximum probability for the $k$th class and uniform values for the others. The $i$th part of the $\bm{c}^{(k)}$ is:
\begin{equation}
    c_i^{(k)} = \left\lbrace
  \begin{array}{@{}ll@{}}
    1-(N-1)p, & \text{if}\ i=k \\
    p, & \text{otherwise,}
  \end{array}\right.
\end{equation}
where $p<\frac{1}{N}$. We want the Aitchison norm of each class composition to be one:
\begin{equation}
  \begin{aligned}
    \forall k \in \{1, \dots N \},~~~~~~\lVert \bm{c}^{(k)} \rVert_a = 1 &\iff \sqrt{\frac{1}{2N} \sum_i^N \sum_j^N \left( \log \frac{c_i^{(k)}}{c_j^{(k)}} \right)^2} = 1,\\
    \text{for clarity, we drop the $(k)$ from the equations,}\\
    &\iff
  \end{aligned}
\end{equation}

\section{Estimation of the Shapley compositions}
\label{app:algo}
In this work, we used an adaptation of Algorithm 2 in \cite{vstrumbelj2014explaining} we present in this section.

@@

When a sample is drawn, the feature for which the sample will be used for improving the Shapley composition estimation is chosen to maximize $\frac{\tr \bm{\Sigma}^{(i)}}{m_i} - \frac{\tr \bm{\Sigma}^{(i)}}{m_i+1}$. Like in \cite{vstrumbelj2014explaining}, this is summarized in Algorithm \ref{alg:2}.
\begin{algorithm}[tb]
   \caption{Adaptation of the Algorithm 1 in \cite{vstrumbelj2014explaining} for approximating the Shapley composition of the $i$th feature, with model $\bm{f}$, instance $\bm{x}\in\mathcal{X}$ and $m$ drawn samples.}
   \label{alg:1}
\begin{algorithmic}
   \STATE Initialize $\bm{\phi_i}\leftarrow \ilr^{-1}(\bm{0})$
   \FOR{$1$ {\bfseries to} $m$}
   \STATE Randomly select a permutation $\pi$ of the set of indexes $\mathcal{I}$,
   \STATE Randomly select a sample $\bm{w}\in\mathcal{X}$,
   \STATE Construct two instances:
   \begin{itemize}
     \item $\bm{b}_1$: which takes the values from $\bm{x}$ for the $i$th feature and the features indexed before $i$ in the order given by $\pi$, and takes the values from $\bm{w}$ otherwise,
     \item $\bm{b}_2$: which takes the values from $\bm{x}$ the features indexed before $i$ in the order given by $\pi$, and takes the values from $\bm{w}$ otherwise.
     \end{itemize}
   \STATE $\bm{\phi}_i \leftarrow \bm{\phi}_i \oplus \bm{f}(\bm{b}_1) \ominus \bm{f}(\bm{b}_2) $
   \ENDFOR
   \STATE $\bm{\phi}_i \leftarrow \frac{\bm{\phi}_i}{m}$
\end{algorithmic}
\end{algorithm}

\begin{algorithm}[tb]
   \caption{Adaptation of the Algorithm 2 in \cite{vstrumbelj2014explaining} for approximating all the Shapley compositions by optimally distributing a maximum number of samples $m_{\text{max}}$ over the $d$ features, with model $\bm{f}$, instance $\bm{x}\in\mathcal{X}$ and $m_{\text{min}}$ the minimum number of samples each feature estimation.}
   \label{alg:2}
   \begin{algorithmic}
     \STATE Initialization: $m_{i} \leftarrow 0$, $\bm{\phi}_i \leftarrow \bm{0}$, $\forall i \in \{1, \dots d\}$,
     \WHILE{$\displaystyle \sum_{i=1}^d m_i < m_{\text{max}}$}
     \IF{$\forall i, m_i \leq m_{\text{min}}$}
     \STATE $j = \underset{i}{\text{argmax}} \left( \frac{\tr \bm{\Sigma}^{(i)}}{m_i} - \frac{\tr \bm{\Sigma}^{(i)}}{m_i+1} \right)$,
     \ELSE
     \STATE pick a $j$ such that $m_j < m_{\text{min}}$,
     \ENDIF
     \STATE $\bm{\phi}_j \leftarrow \bm{\phi}_j +$result of Algorithm \ref{alg:1} for the $j$th feature and $m=1$,
     \STATE update $\bm{\Sigma}^{(j)}$ using an incremental algorithm,
     \STATE $m_j \leftarrow m_j+1$
     \ENDWHILE
     \STATE $\bm{\phi}_i \leftarrow \frac{\bm{\phi}_i}{m_i}$, $\forall i \in \{1, \dots d\}$.
\end{algorithmic}
\end{algorithm}

\section{Correction for the Shapley compositions to be \emph{efficient}}
\label{app:correct}

>>>>

\subsection{The isometric log-ratio transformation}
\label{sec:ilr}

@@

@@

The Shapley compositions can be visualized in the Euclidean space isometric to the simplex thanks to the ILR transformation presented in Section \ref{sec:ilr}. This space has the advantage of being intuitive since it is the standard real $(N-1)$-dimensional vector space we are used too.

@@

In the three classes case, the space is $2$-dimensional. We illustrate this example with the well known Iris classification dataset consisting of a set of flowers descriped by 4 features: sepal length and width and petal length and width. The aim is to predict to which of the three species, setosa, versicolor and virginica, a flower belongs to. In our example we use a Support Vector Machine (SVM) classifier with a radial basis function (rbf) kernel as a classifier. Figure \ref{fig:3classes} shows the explanation of one versicolor instance where \ref{fig:3classesshap} shows the Shapley composition in the ILR space and \ref{fig:3classesshapsum} shows how they move the base distribution to the prediction. Having the highest norm, the petal length is the feature contributing the most on the prediction, moving the base to the versicolor maximum probability decision region. Being orthogonal to the virginica class composition, this suggest that this features does not contribute on the predicted probability for this class. The Shapley composition for the petal width goes straight to the opposite direction of the setosa class vector suggesting that this feature contributes in rejecting this class. The other Shapley compositions have a low norm suggesting these features does not contribute in the prediction.
\begin{figure}
  \centering
  \subfigure[Shapley compositions in the ILR space]{\label{fig:3classesshap}
    \includegraphics[width=0.9\linewidth]{figures/3classes/ilrplot.png}
    }
  \subfigure[Sum of the Shapley compositions in the ILR space from the base to the prediction]{\label{fig:3classesshapsum}
    \includegraphics[width=0.9\linewidth]{figures/3classes/ilrplotsum.png}
  }
  \caption{Shapley explaination in the ILR space for the classification of an Iris instance.}
  \label{fig:3classes}
\end{figure}

@@

In a four classes examples, the simplex is $3$-dimensional. We illustrate this with a simple digit recognition task\footnote{We use the scikit-learn's digits dataset\cite{pedregosa2011scikit}.}. It consists of classifying an $8\times8$ image as representing one of the digit amoung: 0, 1, 2 and 3. Since they are 64 pixels as features which would correspond to 64 Shapley composition, we reduce the number of features to 6 using a principal component analysis for clarity. We again use a SVM classifier with an rbf kernel. The same explaintion analysis as before can be applied here but in a $3$-dimensional plot as illustrated by Figure \ref{fig:4classesshapsum}. To better understand of this space is divided in four regions each representing the maximum probabity region for one class, one can think about the shape of a methane molecule. The hydrogens correspond to the vextex and the carbon to the center of the tetrahedron i.e. a $3$-dimensional simplex. The relative position of the class compositions in the ILR space are the same as the bonds between the corbon and a hydrogen: the angles are $\approx 109.5^{\circ}$.
\begin{figure}
  \centering
  \includegraphics[width=0.9\linewidth]{figures/4classes/ilrplotsum.png}
  \caption{Shapley explaination in the $3$-dimensional ILR space for a four classes digit recognition task. The Shapley compositions are summed in the ILR space from the base distribution to the prediction.}
  \label{fig:4classesshapsum}
\end{figure}

\subsubsection{More classes}
\label{sec:more}
...

\subsection{Summarize explaination with angles, norms and projections}

Some may find the fine analysis of the features contributions in cases with more than four classes, like discussed in Section \ref{sec:more}, tricky. Indeed, this require a careful choose of the sequential binary partition (i.e. the basis) and of the subspace to visualize. However, the Shapley explaination can be summarized by sets of angles, norms and projections.

Indeed the norm of a Shapley composition gives the strength of the contribution the feature in the prediction.

Orthogonality can also be investigated ...
...

\subsection{Histograms}

If one found hard to visualize the proposed Shapley explanation in the ILR space, the Shapley composition can be visualized as histograms like discrete probability distributions.

@@

In this work, the estimation algorithm we used to compute the Shapley compositions is an adaptation of Algorithm 2 in \cite{vstrumbelj2014explaining}. Since the resulting Shapley compositions are approximations, the efficiency property does not necessarily hold without an adjustement. Each Shapley compositions are therefore corrected following a similar method as in the sampling approximation in the SHAP toolkit \cite{NIPS2017_7062}\footnote{\url{https://github.com/shap/shap/blob/master/shap/explainers/_sampling.py}}. See Appendix \ref{app:algo} and \ref{app:correct} for more details.

@@

Compare with standard Shapley ??

We know small expé..., sounds tricky,... first step for a theoretically founded multiclass problems explanations...
...

@@

\newpage
\section{Class compositions}

A $k$-class compositions $\bm{c}^{(k)}  \in \mathcal{S}^N$ is defined as an unit norm composition going straight to the direction of the $k$th class. This is a discrete probability distribution with maximum probability for the $k$th class and uniform values for the others. The $i$th part of the $\bm{c}^{(k)}$ is:
\begin{equation}
    c_i^{(k)} = \left\lbrace
  \begin{array}{@{}ll@{}}
    1-(N-1)p, & \text{if}\ i=k \\
    p, & \text{otherwise,}
  \end{array}\right.
\end{equation}
where $p<\frac{1}{N}$. We want the Aitchison norm of each class composition to be one:
\begin{equation}
  \begin{aligned}
    \forall k \in \{1, \dots N \},~~~~~~\lVert \bm{c}^{(k)} \rVert_a = 1 \iff& \sqrt{\frac{1}{2N} \sum_{i=1}^N \sum_{j=1}^N \left( \log \frac{c_i^{(k)}}{c_j^{(k)}} \right)^2} = 1,\\
    \text{for clarity,}&\\\text{we drop the $(k)$ from the equations,}&\\
    \iff& \sqrt{\frac{1}{2N} \sum_{i=1}^{N}\left( \left(N-1 \right) \left( \log \frac{c_i}{p} \right)^2 + \left( \log \frac{c_i}{1-(N-1)p} \right)^2 \right)} = 1,\\
    \iff& \sqrt{\frac{1}{2N} 2(N-1) \left( \log \frac{p}{1-(N-1)p} \right)^2} =1,\\
    \text{since $p<\frac{1}{N}$ and the norm should be positive:}&\\
    \iff& \sqrt{\frac{N-1}{N}}\log \frac{1-(N-1)p}{p} =1,\\
    \iff& p = \frac{\exp \left( -\sqrt{\frac{N}{N-1}} \right)}{1+ (N-1)\exp \left( -\sqrt{\frac{N}{N-1}} \right )}.
  \end{aligned}
\end{equation}
To summarize, the $i$th part of a $k$-class compositions $\bm{c}^{(k)} \in \mathcal{S}^N$ is given by:
\begin{equation}
    c_i^{(k)} =
  \frac{1}{1+ (N-1)\exp \left( -\sqrt{\frac{N}{N-1}} \right )} \left( \left\lbrace \begin{array}{@{}ll@{}}
     1, & \text{if}\ i=k \\
     \exp \left( -\sqrt{\frac{N}{N-1}} \right), & \text{otherwise,}
  \end{array}\right.\right).
\end{equation}
In this way, $\bm{c}^{(k)}$ is going straight to the direction of class $k$ and uniformly against all the others.

\newpage
\section{Estimation of the Shapley compositions}
\label{app:algo}
In this work, we used an adaptation of Algorithm 2 in \cite{vstrumbelj2014explaining} we present in this section.

@@

When a sample is drawn, the feature for which the sample will be used for improving the Shapley composition estimation is chosen to maximize $\frac{\tr \bm{\Sigma}^{(i)}}{m_i} - \frac{\tr \bm{\Sigma}^{(i)}}{m_i+1}$. Like in \cite{vstrumbelj2014explaining}, this is summarized in Algorithm \ref{alg:2}.
\begin{algorithm}
   \caption{Adaptation of the Algorithm 1 in \cite{vstrumbelj2014explaining} for approximating the Shapley composition of the $i$th feature, with model $\bm{f}$, instance $\bm{x}\in\mathcal{X}$ and $m$ drawn samples.}
   \label{alg:1}
\begin{algorithmic}
   \STATE Initialize $\bm{\phi_i}\leftarrow \ilr^{-1}(\bm{0})$
   \FOR{$1$ {\bfseries to} $m$}
   \STATE Randomly select a permutation $\pi$ of the set of indexes $\mathcal{I}$,
   \STATE Randomly select a sample $\bm{w}\in\mathcal{X}$,
   \STATE Construct two instances:
   \begin{itemize}
     \item $\bm{b}_1$: which takes the values from $\bm{x}$ for the $i$th feature and the features indexed before $i$ in the order given by $\pi$, and takes the values from $\bm{w}$ otherwise,
     \item $\bm{b}_2$: which takes the values from $\bm{x}$ the features indexed before $i$ in the order given by $\pi$, and takes the values from $\bm{w}$ otherwise.
     \end{itemize}
   \STATE $\bm{\phi}_i \leftarrow \bm{\phi}_i \oplus \bm{f}(\bm{b}_1) \ominus \bm{f}(\bm{b}_2) $
   \ENDFOR
   \STATE $\bm{\phi}_i \leftarrow \frac{\bm{\phi}_i}{m}$
\end{algorithmic}
\end{algorithm}

\begin{algorithm}
   \caption{Adaptation of the Algorithm 2 in \cite{vstrumbelj2014explaining} for approximating all the Shapley compositions by optimally distributing a maximum number of samples $m_{\text{max}}$ over the $d$ features, with model $\bm{f}$, instance $\bm{x}\in\mathcal{X}$ and $m_{\text{min}}$ the minimum number of samples each feature estimation.}
   \label{alg:2}
   \begin{algorithmic}
     \STATE Initialization: $m_{i} \leftarrow 0$, $\bm{\phi}_i \leftarrow \bm{0}$, $\forall i \in \{1, \dots d\}$,
     \WHILE{$\displaystyle \sum_{i=1}^d m_i < m_{\text{max}}$}
     \IF{$\forall i, m_i \leq m_{\text{min}}$}
     \STATE $j = \underset{i}{\text{argmax}} \left( \frac{\tr \bm{\Sigma}^{(i)}}{m_i} - \frac{\tr \bm{\Sigma}^{(i)}}{m_i+1} \right)$,
     \ELSE
     \STATE pick a $j$ such that $m_j < m_{\text{min}}$,
     \ENDIF
     \STATE $\bm{\phi}_j \leftarrow \bm{\phi}_j +$result of Algorithm \ref{alg:1} for the $j$th feature and $m=1$,
     \STATE update $\bm{\Sigma}^{(j)}$ using an incremental algorithm,
     \STATE $m_j \leftarrow m_j+1$
     \ENDWHILE
     \STATE $\bm{\phi}_i \leftarrow \frac{\bm{\phi}_i}{m_i}$, $\forall i \in \{1, \dots d\}$.
\end{algorithmic}
\end{algorithm}

\newpage
\section{Adjustement of the estimated Shapley compositions for efficiency}
\label{app:correct}
